### 2014-10-15 AdamRebuttal.txt

Reviewer A:

We want to emphasize why the result about sequential consistency is important.
Here we use "sequential consistency" as a name for the LTS in Figure 3, which
actually completely specifies the observable behavior of a full multiprocessor
system.  Thus, by proving a relationship with that LTS, we are proving what is
perhaps the most natural version of "full functional correctness" in this
domain.

Regarding store atomicity and weak memory models: we define store atomicity as
a property of main memory only, which is quite complex in its own right, with
hierarchies of caches, etc.  There are additional cache-like elements attached
to the individual processors, called by names like "store buffer".  If we
consider these extra buffers as part of the memory, then you're certainly right
that weak memory models don't guarantee atomic stores. However, weak-memory
systems can be built on top of the same store-atomic main memory interface we
give in the paper; it's only the processor-local store buffers that introduce
weaker behavior.

We agree, though, that we should include less speculation on adaptation to weak
memory; and thanks for the bug reports on our discussion of related work.

Reviewer B:

We agree that the underlying LTS proof techniques are not new.  We view our
contribution as more of demonstrating a new application in an area that badly
needs techniques at this level of modularity.

For both Reviewers B and C:

Sorry for not explaining more of our rationale for considering these LTSes as
practical hardware descriptions.  The cache-coherence algorithm, for instance,
is directly transliterated out of [fill in self-contained paper references
here].  We would like to have an automatic translation from Bluespec code, but
we think the two abstraction levels are close enough that it is reasonable to
draw conclusions from our manual translations.

Reviewer C:

Our main theorem uses termination as an arbitrary property that is sufficient
to make the theorem interesting but still easily stated. That's for the same
reason that many problems in computability theory are equivalent to the halting
problem in suitable senses.

Reviewer D:

We apologize for not getting across the sense of the main theorem. Our
optimized architecture actually does fulfill the SC spec, without any side
conditions on programs!  The explanation of why is not trivial, and the proof
carried out throughout the paper justifies the claim.

We also view simplicity of the proof outline (at the level given in the paper)
as a feature, not a bug, of the approach. The actual proof in Coq is much
longer and more detailed (about 10,000 lines).

### 2014-10-15 ArvindRebuttalLatest.txt

We are thankful to the reviewers for their valuable and in-depth comments. We
hope the following clarifications would make them more enthusiastic about
advocating its publication in POPL.

1. Store-atomicity is ubiquitous: Even though systems implement memory models
weaker than sequential consistency (SC), real multiprocessor systems can be
decomposed into multiple parts, one of which is an atomic-memory. In the
simplest form, atomic memory is just a monolithic multiported global memory
(GM), but more commonly, it is implemented using a hierarchy of coherent
caches. Weak memory model behaviors can be understood in terms of local storage
(LS) between each processor and GM. If the processor memory interface is drawn
such that LS is part of GM, then GM will not behave like an atomic store. By
making LS part of the processor model, we assert that all the nuances of the
known weak memory models can be modeled as part of the processor description.
We address the two examples cited by Reviewer A at the end. The proof of
correctness of cache coherence is completely orthogonal to the semantics of the
processor side and hence the memory model issues.

2. Choice of SC: The processor description in Figure 7 implements SC. This
description will have to be modified for weaker memory models, but the
correctness criteria for ROB in Figure 6 will remain the same. For example, one
may not even issue a verification load or issue it sooner. We acknowledge that
more work needs to be done to formally model weaker memory models, including a
treatment of fences.

3. SC performance issues: In the current architecture literature, it's hotly
debated whether weaker memory models can provide better performance than
properly implemented SC. Our SC processor in Figure 7 has high performance
because it is not restricted to sustain only a fixed number of (speculative)
loads in flight. Furthermore, instructions dependent upon these loads can be
processed out of order. A weaker memory model cannot improve upon this
important aspect. The main restriction in our model is that a processor can
issue only one store or one verification load which may be improvable by a
weaker model.

4. LTS for describing and synthesizing hardware:  A Bluespec program is a
transition system and experience has shown that Bluespec descriptions can be
both concise and generate efficient hardware. The cache coherence protocol of
Figure 10 was initially implemented and synthesized into efficient hardware
from Bluespec (Khan et al ISPASS-2012, see also Dave et al Memocode 2006). A
Bluespec program can be transliterated into an LTS in Coq; currently, this is
done manually.

The processor LTS in Figure 7 has an abstract specification of the ROB and
branch predictor with a clearly stated correctness criteria. To derive the
actual hardware, the LTSes for these components have to be specified in much
greater detail. Our ISA is general enough to encompass any CISC ISA including
x86. However the complexity of the detailed LTSes for ROB etc will depend upon
the specific ISA.

5. Reusability of our Cache Coherence proof: MOSI, MOESI, Firefly etc. are very
similar to MSI protocol, though their transition systems will be different. The
way our proof is organized, we will have to prove Lemmas 5 and 6, and
Invariants 2, 3, and 4 for each of these protocols; the rest of the proof will
remain unchanged. In the current version, these lemmas and invariants account
for x% of the proof. Also all these lemmas deal with less than 4 states and
thus can be proven easily using model checking.

------- 
Examples For Reviewer A

For explaining Reviewer A's IRIW behavior, consider a system containing several
in-order cores, each core connected to what we call as ''non-coherent cache''
NCC. All NCCs are connected to a global atomic memory. An NCC can read any
location from memory and store it, and the core can use the data from NCC, if
present. But stores from the core always happen to the global memory, and the
NCC can discard the location that it has, anytime. So, in the example,
initially, NCC of core 4 reads in the location x. Now the sequence of
operations is
Core 1: Wx1 (into global memory)
Core 2: Rx (gets 1 from global memory)
        Ry (gets 0 from global memory, initial value)
Core 3: Wy1 (into global memory)
Core 4: Ry (gets 1 from global memory)
        Rx (gets 0 from NCC)

Similarly, the second example about store buffers can be explained using a
store buffer instead of the NCC, into which stores are written, and loads can
read from. The sequence for this to explain the behavior is:
Core 1: Wx1 (into store buffer, not to memory)
        Ry (read from memory 0)
Core 2: Wy1 (into store buffer, not to memory)
        Rx (read from memory 0)

This global memory is, in reality, implemented using cache coherent protocols.
The separation of coherent caches implementing atomic memory semantics and the
rest of the system is an important contribution of this paper; a version of
this was presented in Arvind (Memory Model = Instruction Reordering + Store
Atomicity, ISCA 2006), and this paper gives a formal proof of this separation,
while extending this idea to speculating systems.

### 2014-10-15 rebuttal.txt

Store-atomicity vs weak memory-models:
We define store-atomicity as a property of just the memory-subsystem, which is
quite complex in its own right, with hierarchies of caches.  There are
additional cache-like elements attached to individual processors, called by
names like "store-buffer".  If we consider these extra buffers as part of the
memory, then Reviewer-A is correct in saying the weak memory-models don't
guarantee atomic stores. However, weak-memory systems are built on top of the
same store-atomic main memory interface defined in LTS of Figure-5; it's only
the processor-local buffers that introduce weaker behavior (we address
Reviewer-A's examples at the end). Proving that the cache-LTS (Figure-10)
preserves store-atomicity (Figure-5) is a major contribution of this paper. We
agree, however, that extension to weaker memory-models needs work.

Choice of SC:
We define "sequential-consistency" as the LTS in Figure-3, which completely
specifies the observable behavior of a full multiprocessor system, including
termination. Thus, by proving a relationship with that LTS, we are proving what
is perhaps the most natural version of "full functional correctness" in this
domain.

Main result (Theorem-10):
We apologize for not getting across the sense of the main theorem. A system
with our speculative processors and store-atomic memory actually does fulfill
the SC spec, without any side conditions on programs or hidden assumptions! We
think the proof is nontrivial. We view simplicity of our proof outline as a
feature, not a bug. The actual proof in Coq is about 10000 lines.

Proof reusability:
MOSI, MOESI, etc. are very similar to MSI protocol. The way our proof is
organized, we will have to just prove Lemmas 5, 6, and Invariants 2, 3, 4 for
each of these protocols. These lemmas deal with less than 4 states and thus can
also be proven easily using model-checking.

LTS specifications for hardware:
Sorry for not explaining our rationale for considering LTSes as practical
hardware descriptions.  The cache-LTS (Figure-10), for instance, is a direct
transliteration from Bluespec-code [Khan]. See also [Dave].  We plan on
automating the translation from Bluespec-code, but we think the two
abstractions are close enough that it is reasonable to draw conclusions from
manual translations.

Processor Specification:
The processor-LTS (Figure-7) is intentionally abstract to capture a superset of
microarchitectures; eg, we do not give the actual transitions for a reorder
buffer implementation but instead give its correctness criteria (Figure-8).
Our ISA is abstract enough to include CISC-instructions and if desired can be
extended with atomic-memory-instructions and memory-barriers.

SC performance:
In the architecture literature, it has been shown that an aggressive
speculative implementation of SC (like ours) may perform as well as weaker
memory-models ([Falsafi], [Hill]). These arguments are orthogonal to
this paper's technical contributions.

Termination-property:
Our main theorem uses termination as an arbitrary property which makes the
theorem interesting but still easily stated. That's for the same reason that
many problems in computability-theory are equivalent to halting-problem in
suitable senses.

-----------------------

Reviewer-C
1. Please note that the buffers between processor and memory-subsystems --
states "ins" and "outs" in Figures 4, 5, 7, 10 -- are multi-capacity buffers.

Reviewer-A
1. Thanks for the bug reports on our related work discussion.

2. Examples:
IRIW behavior:
Consider a system containing several in-order cores, each core connected to
''non-coherent cache'' NCC. All NCCs are connected to a global atomic sotore.
An NCC can read any location from memory and keep it, and the core can load the
location from NCC, if present. But stores from the core always happen to the
global store, and the NCC can discard the location that it has, anytime.

At the beginning, NCC of core 4 reads in the location for x. The following
sequence gives the required behavior:
Core 1: Wx1 (into global memory)
Core 2: Rx (gets 1 from global memory)
        Ry (gets initial value 0 from global memory)
Core 3: Wy1 (into global memory)
Core 4: Ry (gets 1 from global memory)
        Rx (gets 0 from NCC)

Store buffer behavior:
Instead of NCC, each core has a store-buffer into which stores are written, and
loads can read from. The following sequence gives the required behavior:
Core 1: Wx1 (into store buffer, not to memory)
        Ry (gets initial value 0 from global memory)
Core 2: Wy1 (into store buffer, not to memory)
        Rx (gets initial value 0 from global memory)
--------------------------------------------------

Citations in the rebuttal:

[Falsafi] Guiady, C.; Falsafi, B.; Vijaykumar, T. N., "Is SC+ILP=RC?," Computer
Architecture, 1999. Proceedings of the 26th International Symposium on , vol.,
no., pp.162,171, 1999 doi: 10.1109/ISCA.1999.765948 URL:
http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=765948&isnumber=16584

[Hill] Mark D. Hill. 1998. Multiprocessors Should Support Simple
Memory-Consistency Models. Computer 31, 8 (August 1998), 28-34.
DOI=10.1109/2.707614 http://dx.doi.org/10.1109/2.707614

[Khan] Asif Khan, Muralidaran Vijayaraghavan, Silas Boyd-Wickizer, and Arvind.
2012. Fast and cycle-accurate modeling of a multicore processor. In Proceedings
of the 2012 IEEE International Symposium on Performance Analysis of Systems &
Software (ISPASS '12). IEEE Computer Society, Washington, DC, USA, 178-187.
DOI=10.1109/ISPASS.2012.6189224 http://dx.doi.org/10.1109/ISPASS.2012.6189224

[Dave] N. Dave, M. C. Ng, and Arvind. 2005. Automatic synthesis of
cache-coherence protocol processors using Bluespec. In Proceedings of the 2nd
ACM/IEEE International Conference on Formal Methods and Models for Co-Design
(MEMOCODE '05). IEEE Computer Society, Washington, DC, USA, 25-34.
DOI=10.1109/MEMCOD.2005.1487887 http://dx.doi.org/10.1109/MEMCOD.2005.1487887

### 2014-10-15 TransitionsUsefulOld.tex
stuff

We first reduce the proof of store atomicity to several lemmas. We then
present a key insight that we discovered for proving these lemmas in the form
of three fundamental invariants. We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

@@

The lemmas and invariants presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each lemma, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.
%\adamc{or just remind of the source code URL given earlier}

%\subsection{Invariants}
%\label{firstLevel}
Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.

\begin{lemma}
\textit{latestValue}:
At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
$a$ will have the most up-to-date value for that address, \ie{}
\begin{enumerate}
\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
any time $T'$ such that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T \wedge \dt(c,a) = v$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{latestValue}
\end{lemma}

@@

It is pretty intuitive to prove the store atomicity property given
Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
has the required permission, at which point it is guaranteed to have the latest
value, because it is \textit{clean} (as it has no children).

Lemma~\ref{latestValue} requires further decomposition into more
lemmas.

\begin{lemma}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache-tree hierarchy (where
the ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
\textit{compatible} (defined below).
\label{nonAncestorCompatible}
\end{lemma}
\begin{defn}
\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
\label{compatible}
\end{defn}

@@

\begin{lemma}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
$T_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a \textit{clean} cache.
\label{noTransitWrite}
\end{lemma}

\begin{lemma}
\textit{conservative}: The state of an address in a cache is never greater than
the corresponding entry in its parent's directory.
\label{conservative}
\end{lemma}

\begin{lemma}
\textit{localCompatibility1}:
The state of a cache is never less than its directory entry for any of
its children.
\label{localCompatibility1}
\end{lemma}
\begin{lemma}
\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
\textit{compatible}.
\label{localCompatibility2}
\end{lemma}

Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$, which has
two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
the previous paragraph about the relation between the states of a cache and its
ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
Lemma~\ref{nonAncestorCompatible}.

Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
that any \clean{} (node, address) pair has the latest value until time
$T$. Consider a (node, addr) pair at time $T+1$.
If the pair was \clean{} at $T$, then no cache,
save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
store request for that address, thus continuing to be \clean{} and having the latest value.
And if the pair was not \clean{} at time $T$, it will receive data from a cache
that is \clean{} for the (node, address) pair, and no store
request can be processed when the data is in transit (by
Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
continue to have the latest value whenever it is \clean{} for an address at time $T+1$.

\subsection{Key Insight for Proving Lemmas}

@@

Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
either there is no response in flight, only responses from the parent are in
flight, or only responses from the child are in flight. A response from child
downgrades the state, and a response from parent upgrades the directory. Since
the other side simply lags behind, Lemma~\ref{conservative} holds.
Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
just one cache's state for a single address and the corresponding directory
entry, these properties can also potentially be verified using model-checking.
Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
\ref{noCross} and \ref{respFIFO} using model-checking, then
Lemma~\ref{conservative} is already automatically verified for that
protocol.

Analyzing a parent-child pair will suffice in proving
Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
Lemma~\ref{noTransitWrite} is similar to the proof of
Lemma~\ref{nonAncestorCompatible}.

>>>>

We first reduce the proof of store atomicity to several invariants. We then
present a key insight that we discovered for proving these invariants in the form
of three more fundamental invariants. We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy all of our invariants, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

@@

The invariants presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each invariant, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.
%\adamc{or just remind of the source code URL given earlier}

\subsection{Invariants}
\label{firstLevel}
Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.

\begin{inv}
\textit{latestValue}:
At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
$a$ will have the most up-to-date value for that address, \ie{}
\begin{enumerate}
\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
any time $T'$ such that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T \wedge \dt(c,a) = v$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{latestValue}
\end{inv}

@@

It is pretty intuitive to prove the store atomicity property given
Invariant~\ref{latestValue}. A leaf cache can respond to a request only when it
has the required permission, at which point it is guaranteed to have the latest
value, because it is \textit{clean} (as it has no children).

Invariant~\ref{latestValue} requires further decomposition into more
invariants.

\begin{inv}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache-tree hierarchy (where
the ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
\textit{compatible} (defined below).
\label{nonAncestorCompatible}
\end{inv}
\begin{defn}
\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
\label{compatible}
\end{defn}

@@

\begin{inv}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
$T_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a \textit{clean} cache.
\label{noTransitWrite}
\end{inv}

\begin{inv}
\textit{conservative}: The state of an address in a cache is never greater than
the corresponding entry in its parent's directory.
\label{conservative}
\end{inv}

\begin{inv}
\textit{localCompatibility1}:
The state of a cache is never less than its directory entry for any of
its children.
\label{localCompatibility1}
\end{inv}
\begin{inv}
\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
\textit{compatible}.
\label{localCompatibility2}
\end{inv}

Invariants~\ref{conservative} and~\ref{localCompatibility1} imply a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$, which has
two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
$\s(x_2, a)$ are compatible by Invariant~\ref{localCompatibility2}. From
the previous paragraph about the relation between the states of a cache and its
ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
Invariant~\ref{nonAncestorCompatible}.

Using Invariants~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
prove Invariant~\ref{latestValue}. The proof is by induction on time $T$; we assume
that any \clean{} (node, address) pair has the latest value until time
$T$. Consider a (node, addr) pair at time $T+1$.
If the pair was \clean{} at $T$, then no cache,
save itself (because of Invariant~\ref{nonAncestorCompatible}), can process any
store request for that address, thus continuing to be \clean{} and having the latest value.
And if the pair was not \clean{} at time $T$, it will receive data from a cache
that is \clean{} for the (node, address) pair, and no store
request can be processed when the data is in transit (by
Invariant~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
continue to have the latest value whenever it is \clean{} for an address at time $T+1$.

\subsection{Key Insight for Proving Invariants}

@@

Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
\ref{respFIFO}), it is easy to see that Invariant~\ref{conservative} holds.
Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
either there is no response in flight, only responses from the parent are in
flight, or only responses from the child are in flight. A response from child
downgrades the state, and a response from parent upgrades the directory. Since
the other side simply lags behind, Invariant~\ref{conservative} holds.
Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
just one cache's state for a single address and the corresponding directory
entry, these properties can also potentially be verified using model-checking.
Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
\ref{noCross} and \ref{respFIFO} using model-checking, then
Invariant~\ref{conservative} is already automatically verified for that
protocol.

Analyzing a parent-child pair will suffice in proving
Invariants~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
Invariant~\ref{noTransitWrite} is similar to the proof of
Invariant~\ref{nonAncestorCompatible}.
